Active-contours driven registration method for structure-informed segmentation of diffusion MR images

### 2015-04-06 Publications/2015-NeuroImage/00-main.tex
refactoring, add figure of phantoms
my fixes
added new figure
add ISBI reference
major update

\documentclass[5p,authoryear]{elsarticle}
\journal{ArXiv archiving for pre-publication review}

% *** MISC UTILITY PACKAGES ***
\usepackage{glossaries}
\usepackage{booktabs} % to use \toprule
\usepackage{multirow}
\usepackage[colorinlistoftodos]{todonotes}

@@

\providecommand{\nvec}[1]{\hat{\mathbf{#1}}}
\providecommand{\suppl}[1]{\href{http://dx.doi.org/mydoi}{Supplemental Materials}, #1}
\providecommand{\isores}[1]{#1\ensuremath{\times}#1\ensuremath{\times}#1}

@@

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor an-isot-ropy regis-tra-tion iso-tropic Free-Surfer %
align-ed pre-sent}

@@

\begin{document}
\begin{frontmatter}

@@

\title{Active-contours driven registration applied in the correction of
susceptibility-derived distortions of diffusion MR images}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{keyword}
diffusion MRI \sep susceptibility~distortion \sep segmentation %
\sep nonlinear~registration \sep active~contours \sep echo-planar imaging.
\end{keyword}

@@

%\bibliographystyle{plainnat}
% \section*{References}
\bibliographystyle{mystyle}
\bibliography{Remote}

>>>>

\documentclass[5p,authoryear]{elsarticle}
% \documentclass[authoryear,onecolumn]{elsarticle}
\journal{ArXiv archiving for pre-publication review}

% *** MISC UTILITY PACKAGES ***
\usepackage{glossaries}
\usepackage{booktabs} % to use \toprule
\usepackage{multirow}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{xifthen}
\usepackage{xspace}

@@

\providecommand{\nvec}[1]{\hat{\mathbf{#1}}}
\providecommand{\suppl}[1]{\href{http://dx.doi.org/mydoi}{Supplemental Materials}, #1}
\providecommand{\isores}[2][]{#2\ensuremath{\times}#2\ensuremath{\times}#2%
\ifthenelse{\equal{#1}{}}{}{ [#1]}\xspace}

@@

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor an-isot-ropy regis-tra-tion iso-tropic Free-Surfer %
align-ed pre-sent work-flow data-set data-sets}

@@

% http://tex.stackexchange.com/questions/56105/missing-references-heading
\def\bibsection{\section*{References}}
\begin{document}
\begin{frontmatter}

@@

\title{Active-contours driven registration method for structure-informed segmentation of diffusion MR images}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{keyword}
diffusion MRI \sep cortical~parcellation \sep segmentation %
\sep nonlinear~registration \sep active~contours \sep echo-planar imaging.
\end{keyword}

@@

### 2015-04-06 Publications/2015-NeuroImage/03-introduction.tex
major update

The hypothesis under this research is that the registration and segmentation
  problems in \gls*{dmri} can be solved simultaneously, increasing the geometrical
  accuracy of the process.
We propose a tool called \emph{regseg} to exploit the prior information of shapes
  extracted on \gls*{t1} images to register the anatomical reference
  to \gls*{dmri} space, implicitly segmenting these data.
The approach uses active contours without edges \citep{chan_active_2001} that evolve driving a
  free-form deformation field of B-Spline basis functions.
Optimization is performed using a descent strategy of explicit shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}, therefore \emph{regseg}
  does not implement levelsets as most of the presented methods.
Since the nonlinear distortion is aligned with one of the imaging axis (see 
  \autoref{sec:datasets}), \emph{regseg} includes an anisotropic regularizer of
  the displacements field proposed by \cite{nagel_investigation_1986}.
Finally, we evaluate \emph{regseg} using an extension of our instrumentation framework
  \citep{esteban_simulationbased_2014} that simulates known and realistic distortions
  on \gls*{dmri} data.
The evaluation includes a comparison to the \gls*{t2b} correction by integrating in the framework
  an in-house implementation of the method.

>>>>

The hypothesis under this research is that the registration and segmentation
  problems in \gls*{dmri} can be solved simultaneously, increasing the geometrical
  accuracy of the process.
We propose a tool called \emph{regseg} to exploit the prior information of shapes
  extracted on \gls*{t1} images to register the anatomical reference
  to \gls*{dmri} space, implicitly segmenting these data.
The approach uses active contours without edges \citep{chan_active_2001} that evolve driving a
  free-form deformation field of B-Spline basis functions.
Optimization is performed using a descent strategy of explicit shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}, therefore \emph{regseg}
  does not implement levelsets as most of the presented methods.
Since the nonlinear distortion is aligned with one of the imaging axis (see 
  \autoref{sec:human_connectome}), \emph{regseg} includes an anisotropic regularizer of
  the displacements field proposed by \cite{nagel_investigation_1986}.
Finally, we evaluate \emph{regseg} using an extension of our instrumentation framework
  \citep{esteban_simulationbased_2014} that simulates known and realistic distortions
  on \gls*{dmri} data.
The evaluation includes a comparison to the \gls*{t2b} correction by integrating in the framework
  an in-house implementation of the method.

### 2015-04-06 Publications/2015-NeuroImage/04-methods.tex
major update

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
%
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the multispectral target
  volume.
For convenience, and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$:

 	\begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) = \notag\\
  &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
%
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the reference image $M$.
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure01}
	\caption{The active contours are defined as the interfacing surfaces of the competing
	  \glspl{roi} $\Omega_i$.
	They are represented in green and dark blue colors in this close-up.
	They iteratively evolve following their inner normals $\hat{n}_i$ at each vertex
	  $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ are computed as the disparity between data energies of
	  the vertex $\vec{v}_i$ in each limiting region (see \ref{app:shape_priors}).
	In this figure, the gradient speed corresponding to $\vec{v}_1$ is written in the lower
	  box, $\Omega_{wm}$ being the inner limiting region and $\Omega_{gm}$ the outer.
	Finally, every $\bar{s}_i$ is associated with the control points $\vec{u}_k$ through
	  the corresponding weights $\psi_k(\vec{v}_i)$ by Eq. \eqref{eq:gradient_wshape}.
	}\label{fig:method}
\end{figure}

@@

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, + \notag\\
  &+ \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}
%
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The computation of the explicit shape gradients at each $\vec{v}_i'$ is described in \autoref{fig:method}.
The intermediate steps to get to \eqref{eq:final_gradient} from \eqref{eq:gradient_descent} are
  detailed in the \suppl{equations SM6-SM12}:

@@

Finally, to descend this gradient we establish a semi-implicit Euler scheme (see \suppl{section S1.3}),
  with a step size parameter $\delta$, that we solve it in the spectral domain as follows:

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t + \vec{g}_k\} }%
                  {\mathcal{F}\{(\delta^{-1} + \boldsymbol{\alpha})\, I - \boldsymbol{\beta}\Delta\}} \right\}
  \label{eq:update_equation}
  \end{align}

@@

\subsection{Data and corresponding evaluations}
\label{sec:datasets}
In order to assess the performance of \emph{regseg}, we defined a general
  evaluation protocol:
1) Extraction of the set of undistorted surfaces \gammaset{};  
2) computation of a ground-truth field of displacements that is applied to
  generate warped images ($M$) to segment;
3) execution of \emph{regseg} with \gammaset{} and the warped data as inputs; and
4) visual assessment and computation of error metrics.

\paragraph*{Simulated phantoms}\label{sec:digital_phantoms}
The workflow to simulate the digital phantoms and assess
  \emph{regseg} on them is presented in \autoref{fig:evphantoms}.
A set of four binary objects (called ``box'', ``ball'', ``L'',
  and ``gyrus'') are generated combining binarization of
  analytical shapes and mathematical morphology.
The reference surfaces $\gammaset_R$ are extracted from the binary shapes
  using \emph{FreeSurfer} tools.
The ground-truth distortion is generated with a chain of two displacement 
  fields supported by grids of B-Spline basis functions.
The coefficients of the basis functions are generated randomly for
  both levels in their three dimensions.
The three components of the displacements $\vec{u} = (u_d)$ 
  are bounded above by 40\% of the separation between control points
  at each level, to obtain diffeomorphic transforms
  after concatenation \citep{rueckert_diffeomorphic_2006}.
The first deformation field is intended to generate large warpings,
  with control points separated by 50.50mm along the three dimensions
  ($u_d\leq$ 20.20mm).
With the second warping, we look for a resulting field with smoothness
  close to the found in a typical distortion field of \gls*{dmri} data
  \citep{irfanoglu_susceptibility_2011}.
Therefore, the control points are separated 25.25mm ($u_d\leq$ 10.10mm).
Once the ground-truth deformation is generated, the original surfaces
  are warped interpolating the displacements field at each vertex.
The warped surfaces $\gammaset_\text{true}$ are binarized to generate tissue fractions
  in low (\isores[mm]{2.0}) and high (\isores[mm]{1.0}).
Using a \gls*{mr} simulator \citep{caruyer_phantomas_2014}, we synthesize
  \gls*{t1} (TE/TR= 10/1500ms) and \gls*{t2} images (TE/TR= 90/5000ms)
  corresponding to each phantom type, with two resolutions each
  (1.0mm and 2.0mm isotropic).
The field of view at both resolutions is \isores[mm]{100}.
Then, \emph{regseg} is applied to map $\gammaset_R$ into the warped phantoms to
  obtain the registered surfaces ($\hat{\gammaset}_{test}$).
% Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
To quantify the misregistration error, we compute the Hausdorff distance between
 $\hat{\gammaset}_{test}$ and $\gammaset_\text{true}$, using \citep{commandeur_vtk_2011}.

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
% 
For the evaluation of \emph{regseg} on real \gls*{dmri} data of human brains,
  we collected 16 datasets from the \gls*{hcp} database.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing
  pipeline of \emph{FreeSurfer}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large set of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.

\begin{figure}
\includegraphics[width=\linewidth]{figures/figure04-fa-md-jointplot}
\caption{Segmentation model. The prior surfaces $\gammaset_R$ projected into the
  multispectral image to be segmented $M$ actually partition the image in the
  \omegaset{} homogeneous regions.}\label{fig:model}
\end{figure}

The general experimental framework for the real datasets is presented in \autoref{fig:evworkflows}.
\citep{esteban_brain_2012}

We designed the workflow with based on two arbitrary decisions:
First, we map $\gammaset_R$ into a bivariate image comprising the \gls*{fa} and the
  \gls*{adc} maps derived from the distorted \gls*{dmri} dataset.
In a previous study
The underlying segmentation model \omegaset{}
Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  \texttt{mri\_tessellate} to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus, $\Omega_{Tha}$;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures, $\Omega_{VdGM}$;
  3) cerebral \gls*{wm}, $\Omega_{WM}$;
  4) brain stem and cerebellar \gls*{wm}, $\Omega_{bst}$;
	5) cerebellar \gls*{gm}, $\Omega_{cbGM}$; and
	6) cortical \gls*{gm} surface, $\Omega_{pial}$.
The segmentation model corresponding to this selection of surfaces is presented in
  \autoref{fig:model}, and further addressed in \autoref{sec:res_model_and_metric}.

@@

The displacements along the \gls*{pe} axis are related to the local deviation of the
  field $\Delta B_0(\vec{r})$ \citep{jezzard_correction_1995}:

  \begin{equation}
  u(\vec{r}) = \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \: \hat{\vec{e}}_{PE},
  \label{eq:fieldmap}
  \end{equation}
%
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel size along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
With certain \gls*{mr} sequences it is possible to estimate $\Delta B_0$, obtaining
  the so-called fieldmap.

In this case, we derived the deformation $U_{true}$ from the fieldmaps released with
  the corresponding packages of each dataset from the \gls*{hcp} applying \eqref{eq:fieldmap},
  mimicking the real distortions, using a derivation of our previous work
  \citep{esteban_simulationbased_2014}.
After warping the original \gls*{dmri} with $U_{true}^{-1}$, we computed the \gls*{dti} and
  the derived scalars (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.
The stack of \gls*{fa} and \gls*{adc} conforms the multivariate input for \emph{regseg}.

A dual workflow to the general evaluation framework (\autoref{fig:evworkflows})
  was also implemented to integrate the \gls*{t2b} registration scheme.
We reproduced the solution and settings provided with \emph{ExploreDTI}
  \citep{leemans_exploredti_2009}, a widely used toolkit for tractography analysis of
  \gls*{dti}.
\emph{ExploreDTI}, internally uses \emph{elastix} \citep{klein_elastix_2010} to
  perform registration.
First, the average \emph{b0} volume is computed using all the \emph{low-b} volumes within
  the distorted \gls*{dmri} dataset.
Using the ``preprocessed'' instance of the \gls*{t2} image of the corresponding subject as
  reference image, and the \emph{b0} image as moving image, we registered both and
  projected the correct (undistorted) contours to the warped space using the resulting
  transform.

Conversely, the susceptibility-derived distortions only happen along the \gls*{pe}
  axis of the image.
Therefore, a \gls*{swindex} can be computed as the one-to-one distance between corresponding
  vertices of surfaces, weighted by their respective Voronoi area $a_i$:

  \begin{equation}
  sWI = \frac{1}{\sum_i a_i} \sum\limits_i^{N_V} a_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \label{eq:swindex}
  \end{equation}
%
  where $\vec{v}_i \subset \gammaset$ are the locations of the total $N_V$ vertices, and $\hat{\vec{v}}_i$ is the location
  recovered corresponding to the vertex $\vec{v}_i$.

>>>>

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
%
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the moving image.
For convenience, and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$:

 	\begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) = \notag\\
  &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
%
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the image $M$.
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure01}
	\caption{The active contours are defined as the interfacing surfaces of the competing
	  \glspl{roi} $\Omega_i$.
	They are represented in green and dark blue colors in this close-up.
	They iteratively evolve following their inner normals $\hat{n}_i$ at each vertex
	  $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ are computed as the disparity between data energies of
	  the vertex $\vec{v}_i$ in each limiting region (see \suppl{equation SM5})..
	In this figure, the gradient speed corresponding to $\vec{v}_1$ is written in the lower
	  box, $\Omega_{wm}$ being the inner limiting region and $\Omega_{gm}$ the outer, and
    $w_{0,1}$ the relative area associated to vertex $\vec{v}_1$ with respect to
    the total area of surface $\Gamma_0$.
	% Finally, every $\bar{s}_i$ is associated with the control points $\vec{u}_k$ through
	%   the corresponding weights $\psi_k(\vec{v}_i)$ by Eq. \eqref{eq:gradient_wshape}.
	}\label{fig:method}
\end{figure}

@@

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, + \notag\\
  &+ \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\},
  \label{eq:gradient_descent}
  \end{align}
%
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The computation of the explicit shape gradients at each $\vec{v}_i'$ is described in \autoref{fig:method}.
The intermediate steps to get to \eqref{eq:final_gradient} from \eqref{eq:gradient_descent} are
  detailed in the \suppl{equations SM6-SM12}:

@@

Finally, to descend this gradient we establish a semi-implicit Euler scheme (see \suppl{section S1.3}),
  with a step size parameter $\delta$, that we solve in the spectral domain as follows:

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t + \vec{g}_k\} }%
                  {\mathcal{F}\{(\delta^{-1} + \boldsymbol{\alpha})\, I - \boldsymbol{\beta}\Delta\}} \right\}.
  \label{eq:update_equation}
  \end{align}

@@

\subsection{Evaluation protocol}\label{sec:evaluation_protocol}
In order to assess the performance of \emph{regseg}, we defined a general
  evaluation protocol:
1) Extraction of the set of undistorted surfaces $\gammaset_R$;
2) computation of a ground-truth field of displacements $U_{true}$ that is applied to
  generate warped images ($M$) to segment;
3) execution of \emph{regseg} with $\gammaset_R$ and the warped data as inputs; and
4) visual assessment and computation of error metrics.

\subsection{Simulated phantoms}\label{sec:digital_phantoms}
The workflow to simulate the digital phantoms and assess
  \emph{regseg} on them is presented in \autoref{fig:evphantoms}.
A set of four binary objects (called ``box'', ``ball'', ``L'',
  and ``gyrus'') are generated combining binarization of
  analytical shapes and mathematical morphology.
The reference surfaces $\gammaset_R$ are extracted from the binary shapes
  using \emph{FreeSurfer} tools.
The ground-truth distortion is generated with a chain of two displacement 
  fields supported by grids of B-Spline basis functions.
The coefficients of the basis functions are generated randomly for
  both levels in their three dimensions.
The three components of the displacements $\vec{u} = (u_d)$ 
  are bounded above by 40\% of the separation between control points
  at each level, to obtain diffeomorphic transforms
  after concatenation \citep{rueckert_diffeomorphic_2006}.
The first deformation field is intended to generate large warpings,
  with control points separated by 50.50mm along the three dimensions
  ($u_d\leq$ 20.20mm).
With the second warping, we look for a resulting field with smoothness
  close to the found in a typical distortion field of \gls*{dmri} data
  \citep{irfanoglu_susceptibility_2011}.
Therefore, the control points are separated 25.25mm ($u_d\leq$ 10.10mm).
Once the ground-truth deformation is generated, the original surfaces
  are warped interpolating the displacements field at each vertex.
The warped surfaces $\gammaset_\text{true}$ are binarized to generate tissue fractions
  in low (\isores[mm]{2.0}) and high (\isores[mm]{1.0}).
Using a \gls*{mr} simulator \citep{caruyer_phantomas_2014}, we synthesize
  \gls*{t1} (TE/TR= 10/1500ms) and \gls*{t2} images (TE/TR= 90/5000ms)
  corresponding to each phantom type, with two resolutions each
  (1.0mm and 2.0mm isotropic).
The field of view at both resolutions is \isores[mm]{100}.
Then, \emph{regseg} is applied to map $\gammaset_R$ into the warped phantoms to
  obtain the registered surfaces ($\hat{\gammaset}_{test}$).
% Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
To quantify the misregistration error, we compute the Hausdorff distance between
 $\hat{\gammaset}_{test}$ and $\gammaset_\text{true}$, using \citep{commandeur_vtk_2011}.

@@

\subsection{Real datasets} %
\label{sec:human_connectome}
% 
The general experimental framework for the real datasets is presented in \autoref{fig:evworkflows}.
It extends our previous work \citep{esteban_simulationbased_2014} evaluating distortions
  on \gls*{dmri} phantoms.

@@

\paragraph*{Data}
For the evaluation of \emph{regseg} on real \gls*{dmri} data of human brains,
  we collected 16 subjects from the \gls*{hcp} database.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing
  pipeline of \emph{FreeSurfer}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large set of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.

\begin{figure}
\includegraphics[width=\linewidth]{figures/figure04-fa-md-jointplot}
\caption{Segmentation model defined by the homogeneous regions $\Omega_i$ in which
  the image to be segmented is partitioned by the theoretical projection 
  ($\gammaset_{true}$) of $\gammaset_R$ into $M$.
The plot represents a kernel density estimation of the location and spread of
  each region $\Omega_i$ in the bivariate feature space of $M$, comprising 
  the \gls*{fa} and \gls*{adc} maps.
On top and right margins the marginal distributions of $\Omega_i$ are also plotted for
  the \gls*{fa} and \gls*{adc}, respectively.
}\label{fig:model}
\end{figure}

\paragraph*{Segmentation model}
Based on our experience \citep{esteban_brain_2012} and previous studies \citep{ennis_orthogonal_2006},
  we define the moving image as a stack of the \gls*{fa} and \gls*{adc} maps derived
  from \gls*{dmri} data.
After evaluating several alternative models, we empirically defined the partition \omegaset{} 
  of the reference image $R$ as follows:
  1) $\Omega_{Tha}$ including the thalamus;
  2) $\Omega_{VdGM}$ including the ventricular system and deep \gls*{gm} structures;
  3) $\Omega_{WM}$ including the cerebral \gls*{wm};
  4) $\Omega_{bst}$ including the brain stem and cerebellar \gls*{wm};
	5) $\Omega_{cbGM}$ including the cerebellar \gls*{gm}; and
	6) $\Omega_{GM}$ including the cortical \gls*{gm}.
Using tools of \emph{FreeSurfer}, and the appropriate selections of labels in the 
  \emph{aparc} segmentation released with the \gls*{hcp} data, the $\gammaset_R$ set of
  reference surfaces is extracted.
The segmentation model corresponding to this partition is depicted in \autoref{fig:model},
  and discussed with finer detail in \suppl{section S4}.

\paragraph*{Ground-truth generation}
Realistic deformation is achieved generating displacements fields that meet the theoretical
  properties of distortion \citep{jezzard_correction_1995}.
The displacements along the \gls*{pe} axis of the \gls*{dmri} image are related to the local
  deviation of the field $\Delta B_0(\vec{r})$ from its nominal value $B_0$ as follows:

  \begin{equation}
  u_\text{PE} = \frac{\gamma \, T_{acq}\, s_\text{PE}}{2\pi}\Delta B_0(\vec{r})\text{ [mm]},
  \label{eq:fieldmap}
  \end{equation}
%
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time, and
  $s_\text{PE}$ is the pixel size along \gls*{pe}.
Certain \gls*{mr} sequences are designed to estimate $\Delta B_0$, obtaining
  the so-called fieldmap.
We derive the deformation $U_{true}$ from the fieldmap image released with
  the corresponding packages of each dataset from the \gls*{hcp}.
The fieldmap is unwrapped\footnote{fieldmaps are phase maps, clipped in the interval
  of $[-\pi, \pi)$ [rads] or [rads/s].} and smoothed before applying \eqref{eq:fieldmap}.
Then, the original \gls*{dmri} is warped using the resulting displacements field and fed into
  a pipeline that process the corresponding \gls*{dti} and computes the derived scalars of
  interest (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.

\paragraph*{Metric assessment}
Preliminarily, we investigated the aptness of the segmentation model.
For five test datasets, we uniformly sampled the space of distortions
  $\hat{U} = \epsilon \cdot U_{true} = \vec{r} + \epsilon \, u_\text{PE}$
  (with $\epsilon \in [-1.1, 1.1]$ and $u_\text{PE}$ from \eqref{eq:fieldmap}),
  and evaluated the data-term of the cost function \eqref{eq:energy}.
The metric consistently displayed its minimum at the ground-truth location ($\epsilon=0.0$)
  for all the cases (\suppl{Figure S2}).

\paragraph*{Cross-comparison}
A dual workflow to the general evaluation framework of \autoref{fig:evworkflows}
  integrates the alternate \gls*{t2b} registration scheme.
We reproduced the solution and settings provided with \emph{ExploreDTI}
  \citep{leemans_exploredti_2009}, a widely used toolkit for tractography analysis of
  \gls*{dti}.
\emph{ExploreDTI}, internally uses \emph{elastix} \citep{klein_elastix_2010} to
  perform registration.

\paragraph*{Error measurement}\label{sec:experiments_evaluation}
Since the distortions only happen along the \gls*{pe} axis of the image, we compute the
  \gls*{swindex} as the area-weighted distance between corresponding vertices of 
  $\gammaset_{true}$ and their estimation by the method under test $\hat{\gammaset}_{test}$:

  \begin{equation}
  sWI = \frac{1}{\sum_i a_i} \sum\limits_i^{N_V} a_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \label{eq:swindex}
  \end{equation}
%
  where $\vec{v}_i \subset \gammaset_\text{true}$ are the locations of the total $N_V$ vertices,
  and $\hat{\vec{v}}_i \subset \hat{\gammaset}_{test}$ are the locations
  recovered corresponding to $\vec{v}_i$.
In practice, we report the \gls*{swindex} only for three surfaces of interest crucial in whole-brain
  tractography.
These three surfaces delineate the following regions in the model: $\Omega_{VdGM}$, $\Omega_{WM}$ and $\Omega_{GM}$.
